---
title: "MUSA 508 Final Project: Predicting NJ Transit Delays"
author: "Myron Bañez and Amelia Marcantonio-Fields"
output: 
  html_document:
    code_folding: hide
    fig_caption: yes
    toc: true
    toc_float: true
editor_options: 
  markdown: 
    wrap: sentence
  chunk_output_type: console
---
# Introduction + Use Case 

Our firm was contracted by New Jersey Transit to create a model-based dashboard that would aid them in their Capital Improvements Program (CIP) decision making by analyzing systematic delays. Unfortunately, New Jersey Transit has frequent issues with delayed trains, but there are plans to spend more money on infrastructure improvements that will make the system run more smoothly. In 2019, and for the second year in a row, NJT Trains broke down more than any other system in the county (according to the Federal Transit Administration). In January 2021, NJT ordered 25 new dual-powered locomotives that are in the process of being distributed and in 2023, 113 new multi-level rail cars will arrive to replace the 40-year old previous fleet. The mechanical failure rate of the new cars is a fraction of the previous fleet's. 
Our model-based dashboard sets out to help NJT track where delays are the longest and prioritize lines that are experiencing frequent delays with the most profit potential by predicting delays into the future for CIP planning. Currently, the model focuses on the 5 lines that go directly into NYC. Check out the further report and model below: 

## Project Pitch + Dashboard Mockups
See our project pitch video at this link:
The user of this dashboard will hopefully be a NJT employee who works in logistics and does capital improvements planning. Our project pitch includes 3 wireframes of our impending dashboard. The first wireframe is an interactive map that will be colored by delay times. The NJT users will be able to select the line and timeframe to see which lines are struggling most with delays. The second wireframe will present summarized tabular data, where a NJT employee can explore the line, time, avg delay, and ridership. The third and last wireframe is a CIP planning calculator. Once inputting relevant information like the line, timeframe, and intensity of repair, the dashboard will deduce costs spent, but also costs gained from consistent and solid service.

## Source
https://www.nj.com/news/2021/01/nj-transit-trains-ranked-the-worst-in-the-nation-again.html

# Set Up
Below is the setup for the analysis, including packages necessary and color palettes.
```{r setup, results = 'hide', message=FALSE, warning=FALSE, cache=FALSE, echo = FALSE}
options(knitr.table.format = 'markdown')
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, messages=FALSE, include=TRUE, fig.keep = 'all')

library(tidyverse)
library(sf)
library(lubridate)
library(tigris)
library(gganimate)
library(riem)
library(gridExtra)
library(knitr)
library(kableExtra)
library(mapview)
library(tidycensus)
library(ggcorrplot)
library(RColorBrewer)
library(stargazer)

options(tigris_class = "sf")
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")
root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"


palette5 <- c("#eff3ff","#bdd7e7","#6baed6","#3182bd","#08519c")
palette4 <- c("#D2FBD4","#92BCAB","#527D82","#123F5A")
palette2 <- c("#6baed6","#08519c")
palettechloro <-c("#fef0d9", "#fdcc8a", "#fc8d59", "#e34a33", "#b30000")
```

# Importing + Filtering the Data
The code chunk below imports the New Jersey Transit data and completes the initial, necessary cleaning. Data from September and October 2019 is used to train on. These two months were chosen under the pretense that they would illustrate regular usage patterns. November 2019 data will be predicted on. NJT data is filtered for lines going directly into New York City, for which there are 5. Time intervals, especially lines running every hour, are also created for later analysis. 
```{r Import Data}
knitr::opts_chunk$set(echo = T,
                      results = "hide")

setwd("C:/Users/ammar/OneDrive/Desktop/CPLN592/Final/MUSA_508_Final")

Sep19 <- read_csv("September2019.csv") #training data

Oct19 <- read_csv("October2019.csv") #training data

Nov19 <- read_csv("November2019.csv") #data to predict on

AllData <- rbind(Sep19, Oct19, Nov19) #merging them all together

AllData <- #filtering for NJT
  AllData %>%
  filter(type == "NJ Transit" ) 

AllData <- #filtering for lines going directly into NYC
  AllData %>%
  filter(line == "Montclair-Boonton" | line == "Gladstone Branch" | line == "Morristown Line" | line == "Northeast Corrdr" |line == "No Jersey Coast" ) 

# fixing time data
AllData <- #not sure about actual_time #copied from the book but cut out some of the code
  AllData %>% 
  mutate(interval60 = floor_date(ymd_hms(scheduled_time), unit = "hour"),
         interval15 = floor_date(ymd_hms(scheduled_time), unit = "15 mins"),
         week = week(interval60),
         dotw = wday(interval60, label=TRUE)) %>%
  na.omit()
```

## NJT Context Map
Below is a contextual map of the lines selected for prediction analysis. 
```{r Line Data}
NJTRoutes <- 
  st_read("https://opendata.arcgis.com/datasets/e6701817be974795aecc7f7a8cc42f79_0.geojson") %>%
  st_transform('ESRI:102711') %>%
  filter(LINE_CODE == "MC" | LINE_CODE == "GL" | LINE_CODE == "ME" | LINE_CODE == "NE" |LINE_CODE == "NC") 

NJTStations <- 
  st_read("https://opendata.arcgis.com/datasets/4809dada94c542e0beff00600ee930f6_0.geojson") %>%
  st_transform('ESRI:102711')  %>%
  filter(LINE_CODE == "MC" | LINE_CODE == "GL" | LINE_CODE == "ME" | LINE_CODE == "NE" |LINE_CODE == "NC") 

census_api_key("05b9c101eb2ee7dc7abb88140da527ce637ac07f", overwrite = TRUE)

ddACS19_5 <- load_variables(year = 2019, dataset = "acs5", cache = TRUE)

#pulling in total population, # in labor force, # of commuters by commuter rail

vars19 <- 
  c(totalpop = "B01003_001", 
    laborcount = "B23025_002", 
    commuters = "B08301_013")

censustracts <-  
  get_acs(geography = "tract", variables = vars19, 
                year=2019, state=34, output = "wide", geometry=T) %>% 
  st_transform('ESRI:102711')

# Map 
ggplot() + 
  geom_sf(data = censustracts, color = "grey 89", fill = "transparent") +
  geom_sf(data = NJTRoutes, color = "#005DAA") +
  geom_sf(data = NJTStations, color = "#B9C9DF") +
  labs(title = "Chosen NJ Transit Routes",
       caption = "Figure 1") +
  mapTheme(title_size = 11) + theme(legend.position = "none")
```

## Weather Data
The main external variable for this analysis is weather data. Weather has been known to significantly impact transit flow and decision making. The weather attributes pulled for this analysis are temperature, precipitation, and wind-speed. The Newark airport was used for analysis has it is central to the lines chosen. Weather data will be further analyzed throughout the code.
```{r Weather Data}
ymd_hms("2019-10-01 15:48:00")

weather.Data <- 
  riem_measures(station = "EWR", date_start = "2019-09-01", date_end = "2019-11-30")

weather.Panel <-  
  weather.Data %>%
    mutate_if(is.character, list(~replace(as.character(.), is.na(.), "0"))) %>% 
    replace(is.na(.), 0) %>%
    mutate(interval60 = ymd_h(substr(valid, 1, 13))) %>%
    mutate(week = week(interval60),
           dotw = wday(interval60, label=TRUE)) %>%
    group_by(interval60) %>%
    summarize(Temperature = max(tmpf),
              Percipitation = sum(p01i),
              Wind_Speed = max(sknt)) %>%
    mutate(Temperature = ifelse(Temperature == 0, 42, Temperature))
```

# Creating The Ride Panel 
The two biggest lenses of this analysis are spatial and temporal. That being said, a time-series needs to be created for the following analysis. The following chunk of code is the process of creating a ride panel, or an observation of every space/time combinations. Since,  our use case is geared toward long-range CIP planning and long delays, we are interested in having our ride panel relate to "trips" or the total completion of a train line within a certain hour. Therefore, each entry in our ride panel corresponds to how many times a line runs to completion, every hour of our analysis period. Our last step in creating our ride panel is to remove outlier delays, in this case delays that are over 3 hours.
```{r Final Panel}
ride.template <- 
  filter(AllData, week %in% c(35:48))

length(unique(ride.template$interval60)) * length(unique(ride.template$line)) #10920

study.panel <- 
  expand.grid(interval60 = unique(ride.template$interval60), 
              line = unique(ride.template$line))

nrow(study.panel) #10920

trip.count <- 
  ride.template %>%
  group_by(line) %>%
  mutate(max_stops = max(stop_sequence)) #creating a max sequence

ride.panel <- #train id and hour. find max stop sequence 
  trip.count %>%
    mutate(Trip_Counter = 1) %>%
      group_by(interval60, line, train_id, max_stops) %>%
      summarize(Trip_Count = sum(Trip_Counter, na.rm=T),
                delay_minutes = sum(delay_minutes, na.rm=T)) %>%
          mutate(week = week(interval60),
                 dotw = wday(interval60, label = TRUE))

ride.panel <- merge(ride.panel, weather.Panel, by="interval60") #adding all of our data with weather

ride.panel <- #filtering for delays less than 3 hours
  ride.panel %>%
  filter(delay_minutes < 180) 
```

# Feature Engineering 
We now conduct feature engineering to work on the significance of our variables for the model. The feature engineering includes creating delay lags, peak/non peak timing, week day or weekend indicators, and finally weather analysis.
```{r Feature Engineering}
# Time Lag
ride.panel <- 
  ride.panel %>% 
    arrange(line, interval60) %>% 
    group_by(line) %>% 
    mutate(lag1Hours = dplyr::lag(delay_minutes,1), # 1hour
           lag2Hours = dplyr::lag(delay_minutes,2),# 6hour
           lag3Hours = dplyr::lag(delay_minutes,3), # 12hour
           lag6Hours = dplyr::lag(delay_minutes,6), # 24hour
           lag24Hours = dplyr::lag(delay_minutes,24), # 48hour
           lag168Hours = dplyr::lag(delay_minutes,168))  %>%  # 1week
   ungroup()


# Peak 
ride.panel$time <- format(as.POSIXct(ride.panel$interval60), #Getting time column
                    format = "%H:%M:%S")

ride.panel <-
  ride.panel %>% 
  mutate(Peak = case_when(time >= "06:00:00" & time <= "10:00:00" ~ "AM",
                          time >= "16:00:00" & time <= "19:00:00" ~ "PM",
                          TRUE ~ "Non-Peak"))

# Day
ride.panel <-
  ride.panel %>%
  mutate(Day = case_when(
    dotw == "Mon" |dotw == "Tue" | dotw == "Wed" | dotw == "Thu"  | dotw == "Fri" ~ "Weekday",
    dotw == "Sat" |dotw == "Sun" ~ "Weekend"))

# Train ID
ride.panel <-
  ride.panel %>%
  mutate(TrainID = case_when(train_id == "0427" | train_id == "7220" | train_id == "6635" | train_id == "6919" | train_id == "3847" | train_id == "6621" | train_id == "0882" | train_id == "3825" | train_id == "6619" | train_id == "6640" | train_id == "6674" | train_id == "6611" | train_id == "6631" | train_id == "0439" | train_id == "6279" | train_id == "7273" | train_id == "0421" | train_id == "3861" | train_id == "4753" | train_id == "6295" ~ "Offender",
                             TRUE ~ "Non-Offender"))

ride.panel <- 
  ride.panel %>%
  mutate(Offender = case_when(TrainID == "Offender" ~ 1,
                              TrainID == "Non-Offender" ~ 0))

# Weather
ride.panel <-
  ride.panel %>%
  mutate(Weather = case_when(Temperature < 70 ~ "Cold",
                             Temperature >= 70 ~ "Hot"))
```

# Splitting the Data
In order to run the model and conduct predictions, we must split the data. To do this, we assign all data in September and October as our train set and November data as our test set.
```{r Split Training/Test}
ride.Train <- filter(ride.panel, week < 45)
ride.Test <- filter(ride.panel, week >= 45)
```

# Exploratory Analysis
Our feature engineering and model creation was built on the following analysis. This analysis section includes a wealth of visuals relating delays to the various features of the model. 
- Figures 2.1a/b/c chart weather over the 3-month period. 
- Figure 2.2a/b charts lines and delay minutes of data from Sept & Oct 2019
- Figure 2.3 depicts delay minutes per day over each week
- Figure 2.4 shows delay minutes as a function of the spatial lags 
- Figure 2.5 shows delay minutes as a function of temperature
- Figure 2.6 is a histogram of log(delays)
- Figure 2.7 is a correlation plot of numeric variables in the model
- Figure 2.8 depicts delay minutes as a function of max stops (how many stops on a line)
- Figure 2.9 shows delay minutes as a function of various categorical variables

NEED TO MENTION HIGHLIGHTS

```{r Exploratory Analysis}

# charting weather over the 3 month period
grid.arrange(top = "Weather Data - NYC & North Jersey - September to November 2019",
  ggplot(weather.Panel, aes(interval60,Percipitation)) + geom_line() + 
    labs(title="Percipitation", x="Hour", y="Percipitation", caption = "Figure 2.1a") + plotTheme(),
  ggplot(weather.Panel, aes(interval60,Wind_Speed)) + geom_line() + 
    labs(title="Wind Speed", x="Hour", y="Wind Speed", caption = "Figure 2.1b") + plotTheme(),
  ggplot(weather.Panel, aes(interval60,Temperature)) + geom_line() + 
    labs(title="Temperature", x="Hour", y="Temperature", caption = "Figure 2.1c") + plotTheme())

#charting delays by line
ride.Train %>% #This is here just t see delays. we could takeit out
  group_by(line) %>% 
  summarise(Delay = sum(delay_minutes)) %>%  kable(caption = "Lines by Delays (Figure 2.2a)") %>%
  kable_styling("striped",full_width = F) 

group_by(ride.Train, line) %>% #Delay by line
  ggplot(aes(reorder(line, delay_minutes, FUN = max), 
             log(delay_minutes), fill = line)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = palette5) +
    labs(x = "Line", y = "Delay Minutes", title= "Delay minutes by line", 
         caption = "Figure 2.2b") +
    plotTheme() + theme(legend.position = "none") 

# Trip Count Serial Autocorrelation EA
mondays <- 
  mutate(ride.panel,
         monday = ifelse(dotw == "Mon" & hour(interval60) == 1,
                         interval60, 0)) %>%
  filter(monday != 0) 

spooky   <- as.POSIXct("2019-10-31 01:00:00 UTC")

rbind( #change this?
  mutate(ride.Train, Legend = "Training"), 
  mutate(ride.Test, Legend = "Testing")) %>%
    group_by(Legend, interval60) %>% 
      summarize(DM = sum(delay_minutes)) %>%
      ungroup() %>% 
      ggplot(aes(interval60, DM, colour = Legend)) + geom_line() +
        scale_colour_manual(values = palette2) +
        geom_vline(xintercept = spooky, linetype = "dotted") +
        geom_vline(data = mondays, aes(xintercept = monday)) +
        labs(title="Delay minutes by week: September-November 2019",
             subtitle="Dotted line for Halloween", 
             x="Day", y="Delay Minutes", caption = "Figure 2.3") +
        plotTheme() + theme(panel.grid.major = element_blank()) 

# Time Lags
plotData.lag <-
  filter(as.data.frame(ride.panel), week == 42) %>%
  dplyr::select(starts_with("lag"), delay_minutes) %>%
  gather(Variable, Value, -delay_minutes) %>%
  mutate(Variable = fct_relevel(Variable, "lag1Hour","lag2Hours","lag3Hours", "lag6hours", "lag24hours", "lag168hours"))

correlation.lag <-
  group_by(plotData.lag, Variable) %>%
    summarize(correlation = round(cor(Value, delay_minutes, use = "complete.obs"), 2)) 

ggplot(plotData.lag, aes(Value, delay_minutes)) +
  geom_point(size = 0.1) +
  geom_text(data = correlation.lag, aes(label = paste("r =", round(correlation, 2))),
            x=-Inf, y=Inf, vjust = 1.5, hjust = -.1) +
  geom_smooth(method = "lm", se = FALSE, colour = "blue") +
  facet_wrap(~Variable, ncol = 3, scales = "free") +
  labs(title = "Delay minutes as a Function of Time Lag",
       subtitle="One week in October 2019", 
       caption = "Figure 2.4") +
  plotTheme()

# Weather EA
ride.panel %>%
  group_by(interval60) %>% 
  summarize(DM = mean(delay_minutes),
            Temperature = first(Temperature)) %>%
  mutate(week = week(interval60)) %>%
  ggplot(aes(Temperature, DM)) + 
    geom_point() + geom_smooth(method = "lm", se= FALSE) +
    facet_wrap(~week, ncol=3) + 
    labs(title="Delay minutes as a fuction of Temperature by week",
         x="Temperature", y="Mean Trip Count", 
         caption = "Figure 2.5") +
    plotTheme()

# Bar Plot
options(scipen=999)

ggplot(ride.panel, aes(x = log(delay_minutes))) + labs(y = "Count", x = "Log(Delay Minutes)", title = "Histogram of Delays", caption = "Figure 2.6") + 
    geom_histogram()
  
# Correlation Matrix  
numericVars <- 
  select_if(ride.panel, is.numeric) %>% na.omit()

ggcorrplot(
  round(cor(numericVars), 1), 
  p.mat = cor_pmat(numericVars),
  colors = c("#25CB10", "white", "#FA7800"),
  type="lower",
  insig = "blank") +  
    labs(title = "Correlation Across Numeric Variables", caption = "Figure 2.7") 

# Extra
ride.panel %>% 
  dplyr::select(delay_minutes, max_stops) %>% 
  gather(Variable, Value, -delay_minutes) %>% 
   ggplot(aes(Value, delay_minutes)) +
     geom_point(size = .5) + geom_smooth(method = "lm", se=F, colour = "#FA7800") +
     facet_wrap(~Variable, ncol = 3, scales = "free") +
     labs(title = "Delay minutes as a function of max stops", 
          caption="Figure 2.8") +
     plotTheme()

ride.panel %>%
  dplyr::select(delay_minutes, dotw, Peak, Day, Weather) %>%
  gather(Variable, Value, -delay_minutes) %>% 
   ggplot(aes(Value, delay_minutes)) +
     geom_bar(position = "dodge", stat = "summary", fun.y = "mean") +
     facet_wrap(~Variable, ncol = 1, scales = "free") +
     labs(title = "Delay minutes as a Function of\nCategorical Variables", y = "Delay Minutes", 
          caption="Figure 2.9") +
     plotTheme() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Regression Models 
The code below are the 5 iterations of our predictive models. Regression 1 is the baseline regression with no feature engineered or overly descriptive variables included. Regression 4 is our best suited model and includes all feature engineered variables.

```{r Linear Regression Models}
# Linear - Baseline
test.reg.linear1 <- lm(log(delay_minutes) ~ hour(interval60) + dotw + line, data=ride.Train %>% mutate(delay_minutes = delay_minutes + 0000.1)) # STARGAZER HERE!!
summary(test.reg.linear1) #rsquare .031

# Linear - Feature Engineered
test.reg.linear2 <- lm(log(delay_minutes) ~ hour(interval60) + dotw + line + lag1Hours + lag2Hours +lag3Hours + lag6Hours + lag24Hours + lag168Hours, data=ride.Train %>% mutate(delay_minutes = delay_minutes + 0000.1))
summary(test.reg.linear2) #rsqaure .033

test.reg.linear3 <- lm(log(delay_minutes) ~ hour(interval60) + dotw + line + train_id + lag1Hours + lag2Hours +lag3Hours + lag6Hours + lag24Hours + lag168Hours + Day + Peak + Weather, data=ride.Train %>% mutate(delay_minutes = delay_minutes + 0000.1))
summary(test.reg.linear3) #rsqaure .32

test.reg.linear4 <- lm(log(delay_minutes) ~ hour(interval60) + dotw + (line * Peak) + Offender + lag1Hours + lag2Hours +lag3Hours + lag6Hours + lag24Hours + lag168Hours + Temperature + Wind_Speed + Percipitation + Trip_Count, data=ride.Train %>% mutate(delay_minutes = delay_minutes + 0000.1)) # STARGAZER HERE!!
summary(test.reg.linear4) #rsquare .34

test.reg.linear5 <- lm(log(delay_minutes) ~ hour(interval60) + dotw + (line * Peak) + lag1Hours + lag2Hours +lag3Hours + lag6Hours + lag24Hours + lag168Hours + Weather + Trip_Count, data = ride.Train %>% mutate(delay_minutes = delay_minutes + 0000.1))
summary(test.reg.linear5) #rsquare .34
```

## Regression Model Summaries
```{r model sum stats, results='asis'}
#---Polished LM table of training dataset
stargazer(test.reg.linear4, type="html", digits=1, title="Linear Model of Training Dataset (Figure 3)", out = "Training LM.txt")
```

# Validation

## Validation by Regression
Below are regression validations by time (both delay minutes and weeks). It is clear in these models that our featured engineered, or "best regression," predicts the best. The overall error of the best regression is a 16 minute MAE. Figure 4.2 depicts how much better the best regression predicts and follows the variation of the observed delays much better.  
```{r Validation by Regression}
# Validation by Regression
ride.Test.weekNest <- 
  as.data.frame(ride.Test) %>%
  nest(-week) 

model_pred <- function(dat, fit)
  {
   pred <- exp(predict(fit, newdata = dat))
}

week_predictions <- 
  ride.Test.weekNest %>% 
    mutate(BaselineRegression = map(.x = data, fit = test.reg.linear1, .f = model_pred),
           BestRegression = map(.x = data, fit = test.reg.linear4, .f = model_pred))

week_predictions <- week_predictions %>%  
    gather(Regression, Prediction, -data, -week) %>% 
    mutate(Observed = map(data, pull, delay_minutes),
           Absolute_Error = map2(Observed, Prediction,  ~ abs(.x - .y)),
           MAE = map_dbl(Absolute_Error, mean),
           sd_AE = map_dbl(Absolute_Error, sd))

week_predictions %>%
  dplyr::select(week, Regression, MAE) %>%
  gather(Variable, MAE, -Regression, -week) %>%
  ggplot(aes(week, MAE)) + 
    geom_bar(aes(fill = Regression), position = "dodge", stat="identity") +
    scale_fill_manual(values = palette5) +
    labs(title = "Mean Absolute Errors by model specification and week", caption = "Figure 4.1") +
  plotTheme()

week_predictions %>% 
  mutate(interval60 = map(data, pull, interval60)) %>%
  dplyr::select(interval60, Observed, Prediction, Regression) %>%
  unnest() %>%
  gather(Variable, Value, -Regression, -interval60) %>%
    group_by(Regression, Variable, interval60) %>%
    summarize(Value = mean(Value)) %>%
    ggplot(aes(interval60, Value, colour=Variable)) + geom_line(size = 1.1) + 
      facet_wrap(~Regression, ncol=1) +
      scale_colour_manual(values = palette2) +
      labs(title = "Mean Predicted/Observed delay minutes by hourly interval", 
           x = "Hour", y= "Delay Minutes", caption = "Figure 4.2") +
  plotTheme()
```

## Validation By Line


```{r Validation by Line}
# Validation by Line
wk45 <- unnest(week_predictions[5,])
wk46 <- unnest(week_predictions[6,])
wk47 <- unnest(week_predictions[7,])
wk48 <- unnest(week_predictions[8,])

test <- rbind(wk45, wk46, wk47, wk48)

ggplot(test, aes(x=MAE)) + 
  geom_histogram(fill = "blue") +
  facet_wrap(~line, ncol=1) +
  labs(title = "Cross Validation Tests in Mean Average Error", caption = "Figure 5.1") +
  plotTheme()

test %>%  
  gather(Variable, Value, Prediction, Observed, -line, -interval60) %>%
    group_by(line, Variable, interval60) %>%
    summarize(Value = mean(Value)) %>%
    ggplot(aes(interval60, Value, colour=Variable)) + geom_line(size = 1.1) + 
      facet_wrap(~line, ncol=1) +
      scale_colour_manual(values = palette2) +
      labs(title = "Mean Predicted/Observed Delay Minutes", 
           subtitle = "By line by hourly interval",
           x = "Hour", y= "Delay Minutes", caption = "Figure 5.2") +
      plotTheme()

test %>% 
  group_by(line) %>%
  summarize(MAE = mean(MAE, na.rm = T),
            AvgDelay = mean(delay_minutes, na.rm=T)) %>%
  kable(title = "MAE and average delay time by Line", caption = "Figure 5.3") %>%
  kable_styling("striped",full_width = F)

test %>% # TRAIN ID
  group_by(line, train_id) %>%
  summarize(AvgDelay = mean(delay_minutes, na.rm=T)) %>%
  filter(AvgDelay > 58.5) %>%
  kable(title = "Train ID and average delay time by Line", caption = "Figure 5.4") %>% kable_styling("striped",full_width = F)

#Adding delay during peak time 
test_peakdelay <- 
  test %>% 
  group_by(line) %>%  
  filter(Peak == "AM" | Peak == "PM")

test_AMdelay <- # trying to get peak in the previous line was HARD, so i filtered lol
  test %>%
  group_by(line) %>%
  filter(Peak == "AM")

test_AMdelay <- # gives you predicted avg MORNING delay time during peak by line
test_AMdelay %>% 
  group_by(line) %>%
  mutate(MorningDelay = mean(Prediction))

test_PMdelay <- # repeated the process for the evening
  test %>%
  group_by(line) %>%
  filter(Peak == "PM")

test_PMdelay <- # predidcted EVENING delay time during peak by line
test_PMdelay %>% 
  group_by(line) %>%
  mutate(EveningDelay = mean(Prediction))

test_delays <- rbind(test_AMdelay, test_PMdelay) # brought them bitches together

test_delays %>% 
  group_by(line) %>%
  summarize(MorningDelay = mean(MorningDelay, na.rm=T),
            EveningDelay = mean(EveningDelay, na.rm=T)) %>%
  kable(title = "Average delays during peak time by Line", caption = "Figure 5.5") %>%
  kable_styling("striped",full_width = F)

group_by(test, line) %>% #Delay by line
  ggplot(aes(reorder(line, Prediction, FUN = max), 
             delay_minutes, fill = line)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values = palette5) +
    labs(x = "Line", y = "Delay Minutes", title= "Predicted Delay minutes by line", caption = "Figure 5.6") +
    plotTheme() + theme(legend.position = "none") 
```

# Final Context Map

```{r final context map}
NovPrediction <- test %>%
  group_by(line) %>%
  summarize(totaldelay = sum(Prediction))

RoutesPredictions <- cbind(NJTRoutes, NovPrediction)

ggplot() + 
  geom_sf(data = censustracts, color = "grey 89", fill = "transparent") +
  geom_sf(data = RoutesPredictions, aes(color = as.factor(totaldelay)), show.legend = "delay times") +
  scale_color_manual(name = "line", values = palettechloro) +
  labs(title = "Chosen NJ Transit Routes: November 2019 Predicted Delays",
       caption = "Figure 6.1", fill = "line") +
  mapTheme(title_size = 11) + theme(legend.position = "right")
```


# Conclusion 

## Accuracy + Generalizability 
 
## How to Improve 

## Policy Recommendations